---
title: "Iteration in R"
author: "Lionel Henry"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Iteration in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library("flowery")
```


## Vector operations versus chunked data flow

Every piece of data is a vector in R. Lists and data frames are finite collections of atomic vectors and hold all of their data in memory.  Accordingly, R programming is vector-oriented. Arithmetic operations like `+` are vectorised and applying functions over vectors (e.g. with `base::lapply()` or `purrr::map()` and variants) is a favourite of R programmers.

However the vector approach only works when all of the data fits in memory. If it doesn't fit, the data needs to be processed chunk by chunk. This is a task that is not well structured in R as of now. The purpose of flowery iterators is to provide the tools and idioms needed for structuring the generation and transformation of streams of chunked data.

The flowery package provides three complementary features:

  * **Iterators**: An iterator is a reference to the current data chunk that has two basic operations: advancing to the next chunk with `advance()` and dereferencing the current chunk with `deref()`. Advancing causes the actual generation of the next data chunk. You can use regular functions to create an iterator but it is often easier to use generators.

  * **Generators**: Generators are the preferred way of creating iterators in flowery. They are special functions that can *pause* themselves. This feature is particularly handy for creating chunks of data iteratively.

  * **Transformation steps**: Also called [transducers](https://clojure.org/reference/transducers), transformation steps make it easy to adapt iterators by mapping a function over incoming chunks or discarding unwanted elements.  While understanding how transformation steps work under the hood is a bit involved, using them on iterators is very easy.

This vignette explores how flowery envisions chunked iteration in R.  It describes what it takes to generate data chunkwise, how the iterator wrappers structure the process of iterating over chunks, and finally how generators provide syntax that makes the process easy.


## Iteration

Iteration is a stateful thing. It is about advancing from one state to the next. Consider that simple loop:

```{r}
greet <- function(x) paste("hey", x)
x <- character(length(letters))

for (i in seq_along(letters)) {
  x[[i]] <- greet(letters[[i]])
}

x
```

At each step the state changes. In this example the state is pretty simple, it is the index variable `i`. This changing state lets us *generate* different pieces of data. In this case `letters[[i]]` generates simple strings containing the letters of the alphabet. In a more involved case, we might pull out heavy data frames from a database.

What if we wanted to reuse that piece of code in a function with the objective of returning the data one piece at a time? Simply wrapping our loop only works when we want to return the whole vector:

```{r}
iter <- function(x) {
  x <- character(length(letters))

  for (i in seq_along(letters)) {
    x[[i]] <- greet(letters[[i]])
  }

  x
}
iter(letters)
```

Instead we need to emulate the loop manually. This means that we need a placeholder for the state! One way to achieve this is with function factories. See the relevant [Advanced R chapter](http://adv-r.had.co.nz/Functional-programming.html) if you are not familiar with factories. Basically, a factory is a function that returns another function. Any data that you create there can be accessed by the new function. That's a great way of keeping state between invocation:

```{r}
make_iter <- function(x) {
  # Create placeholder for iteration state
  i <- 0

  # Create and return the new iterator function
  function() {
    # Use the `<<-` operator to update non-local variables
    i <<- i + 1
    greet(x[[i]])
  }
}
iter <- make_iter(letters)
```

Now let's call our new iterator function:

```{r}
iter(); iter()
```

To sum up, creating an iterator with a function factory involves two phases:

- Creating a persistent state in the environment of the returned iterator.

- Updating this state from the iterator function with the `<<-` operator.

We'll see later that generators make it very easy to manage the iteration state. But first let's see what is missing from the iterator that we have created in order to use it effectively.


## Iterator operations

We have created a function that generates chunks of data on demand.  We need to apply three operations to work with this iterator effectively:

- Advance to a new chunk.
- Get the current chunk.
- Check if the iterator is done.

The first two operations are accomplished in a single step simply by calling the iterator function:

```{r}
iter <- make_iter("foo")

# We advance the iterator and obtain the next value in one call:
x <- iter()
x
```

However how can we check that we are done iterating? The iterator that we created had only one element to iterate over. If we call it again we get an out-of-bounds error:

```{r, error = TRUE}
iter()
```

We need some kind of convention to signal termination of an iterator.  In Python you would do this by throwing a special kind of exception.  That requires wrapping iterators with `tryCatch()` each time they are advanced. For flowery we decided to signal termination with a sentinel value instead. The `NULL` value is a natural candidate. It is especially practical for iteration because many control flow operators return `NULL`. Let's modify our iterator factory so that it returns `NULL` when it is done:

```{r}
make_iter <- function(x) {
  i <- 0
  n <- length(x)  # We'll need to check the length of the vector

  function() {
    i <<- i + 1

    # Return NULL if we are out of bounds
    if (i <= n) {
      greet(x[[i]])
    } else {
      NULL
    }
  }
}

iter <- make_iter("foo")
iter(); iter()
```

Not bad. Now how can we use this sentinel value to stop iterating at the right time? One way to do it would be to assign the next value in a temporary object and check that this value is not the sentinel. We can do both at the same time in a loop condition:

```{r}
iter <- make_iter(c("foo", "bar"))

while (!is.null(elt <- iter())) {
  print(elt)
}
```

This is a bit hard to read though. For this reason flowery provides an iterator constructor that provides structured operations.


## Flowery iterators

A flowery iterator is a wrapper around an iterable function. An iterable function is a function that:

- can be called without arguments to get the next value;
- returns NULL when it is done.

Luckily our iterator factory already meets both these specifications!  We can supply the iterable functions to `new_iterator()` to create a proper flowery iterator:

```{r}
make_flowery_iter <- function(x) new_iterator(make_iter(x))
```

The iterator works just like before:

```{r}
iter <- make_flowery_iter(letters[1:8])
iter(); iter()
```

But you can now use `advance()` and `deref()`. `advance()` returns `TRUE` or `FALSE` depending on whether the next value was a termination sentinel. `deref()` then dereferences the iterator to get its current value:

```{r}
advance(iter)
deref(iter)
```

This is handy in loops:

```{r}
while (advance(iter)) print(deref(iter))
```

A flowery iterator verbosely fails if you try to reenter it when it has exhausted its values:

```{r, error = TRUE}
is_done(iter)
iter()
```

Flowery iterators are terminated with `NULL` but in some cases it might make sense to return `NULL` literally. In this case you can return a `null_box()`. It is automatically transformed to a `NULL` value and does not cause the iterator to terminate. In the following example we create an infinite iterator that always returns `NULL` without ever terminating:

```{r}
iter <- new_iterator(function() null_box())

iter()
iter()
```

Finally you can use `drain()` and its variants to materialise the iterator into a vector. These functions sink all iterated values into a vector. `drain()` creates a list:

```{r}
iter <- make_flowery_iter(letters[1:3])
drain(iter)
```

While the other typed variants create atomic vectors:

```{r}
iter <- make_flowery_iter(letters[1:3])
drain_chr(iter)
```

It is perfectly possible for an iterator to never terminate. In this case you might want to take only a specific number of elements:

```{r}
iter <- make_flowery_iter(letters)
take_chr(iter, 5)
```


## Generators

The purpose of generators is to maintain state. This is why they are ideally suited for generation of data during iteration. The iterators that we created manually in the previous section can easily be turned into a generator:

```{r}
iter <- generator({
  for (x in letters) yield(greet(x))
})
```

The generator can be used just like any iterator:

```{r}
iter(); iter()
```

There are two things going on:

- `generator()` returns an iterator function.

- `yield()` pauses the iterator function. When it is reentered, execution resumes from there. In this case the `for` loop inside our generator continues iterating as if the generator never exited.

When a generator has yielded, it can be called or advanced to the next value and its execution will resume right at the yielding point. The ability to pause and resume considerably simplifies the task of generating data chunk by chunk because the state of the iterator is kept between iteration steps.

Note that generators (and iterators in general) are one-shot. Once iteration is done, reentering the generator issues an error. Let's drain the remaining elements from the iterator:

```{r}
drain_chr(iter)
```

From that point on, reentering the iterator is an error:

```{r, error = TRUE}
iter()
```

For this reason it is often a good idea to provide a **generator factory**. The principle is identical to function factories except that we don't have to maintain the iteration state. Creating a generator factory is thus straightforward:

```{r}
make_iter <- function(x) {
  generator({
    for (elt in x) yield(greet(elt))
  })
}

iter <- make_iter(c("foo", "bar"))
drain(iter)
```

This generator factory can be reused as many times as needed to create fresh generators:

```{r}
iter <- make_iter(c("baz", "barbaz"))
drain(iter)
```

Note finally that generators don't have to yield, they can also return. Once it has returned the generator is done:

```{r}
iter <- generator({
  x <- 1:10

  for (elt in x) {
    if (elt < 5) {
      yield(elt)
    } else {
      return(100)
    }
  }
})

drain_dbl(iter)
```


## Transforming iterators

As we have seen iterator and generators are useful for generating data chunk by chunk in a structured way. The flowery package also offers several ways of transforming the stream of data chunks.


### Chaining generator expressions

Inside a generator you can supply iterators to `for` loops. This makes it easy to chain generators:

```{r}
integer_iter <- generator(for (x in 1:10) yield(x))
square_iter <- generator(for (x in integer_iter) yield(x^2))
```

in addition a shorthand `gen()` alias is provided for `generator()`.  There is no difference between the two versions. The long version is appropriate in generator factories while the shorthand reads better in generator expressions:

```{r}
integer_iter <- as_iterator(1:10)
square_iter <- gen(for (x in integer_iter) yield(x^2))
final_iter <- gen(for (x in square_iter) yield(x * 100))
```

Draining the final iterator iterates over all iterators in turn. What happens to intermediate values? They are discarded from memory as iteration advances. Only the final summary value is kept in the final output vector:

```{r}
drain_dbl(final_iter)
```


### Transformation step

Transformation steps bring the familiar `map`, `keep` or `discard` idioms that you know from purrr to iterators. You can map a function over iterated chunks with `iter_map()` and discard or keep unwanted elements with `iter_discard()` and `iter_keep()`. You apply these steps to iterators with `iter_adapt()`, which takes an iterator and steps as inputs and returns another iterator.

Let's apply steps so that odd numbers are discarded and all values that pass through the filter are squared by mapping the `^` operator. Note that transformation steps support purrr's `~` notation for lambda functions:

```{r}
iter <- as_iterator(1:10)

iter <- iter_adapt(iter,
  iter_discard(~ . %% 2 != 0),
  iter_map(`^`, 2)
)

drain_dbl(iter)
```


### Iteration with for loops

Finally, flowery provides `iterate()`. It takes a `for` loop expression and instruments it so it understands flowery iterators.

```{r}
iter <- gen(for (x in 1:10) yield(x^2))

iterate(for (x in iter) {
  cat("iteration:", x, "\n")
})
```
